Constructing a universe for the setoid model

### 2021-01-21 paper.tex
Expand apostrophic abbreviations: * it's -> it is * can't -> cannot * wasn't -> was not * ...

For convenience, we slightly deviate from $\mlttp$ both in the paper and in the
formalization, for instance by relying on pattern matching instead of
eliminators, and using primitive versions of $\mProp$-valued $\Pi$ and $\Sigma$
types instead of deriving them from truncation. We operate under the assumption
that everything can be equivalently carried out in $\mlttp$, although we haven't
fully checked all the necessary details.

@@

We can show that the setoid model validates the usual basic type formers ($\Pi,
\Sigma$, etc.), function extensionality and a universe of strict
propositions with propositional extensionality \cite{setoid99}. Note that
we don't need identity types or inductive types (W-types) for this.

@@

Other design choices on the setoid universe are less essential, but still
require careful consideration. For instance, one question is whether the setoid
universe should support universe induction, thus exposing the inductive
structure of the codes. Such an elimination principle is known to be inconsistent
with univalence, although this is not an issue in our case; nevertheless it's
not immediately clear if the elimination principle can be justified by the
semantics, that is, if our encoding of the setoid universe in the model allows
to define such a universe eliminator.
%
The question arises because our final encoding of the setoid universe only
supports a weak form of elimination, for reasons that are explained in
Section~\ref{inductive-universe}. Although not currently needed, a stronger
eliminator might be necessary to justify universe induction. This problem should
not arise in the other encodings of the setoid universe (as given in
Section~\ref{ir-setoid-universe} and Section~\ref{ii-universe}).

@@

In the next section we transform our current inductive-recursive definition to
one that doesn't use induction-recursion.
%
The way this is done is inspired by the well-known trick to eliminate
induction-recursion described in Section~\ref{ir-universes}, but modified in a
novel way to account for the presence of $\mProp$-valued types. To our
knowledge, this is the first time this reduction method is applied to an
inductive-recursive type of this kind.

@@

Unfortunately this definition only manages to capture the idea, but does not
actually typecheck. In fact, $\_\sim_{\setoidU}\_$ should be a $\mProp_1$-valued
relation, so $A \sim_{\setoidU} B$ should be a proposition. However, the
$\Sigma$ type shown above clearly isn't, since it quantifies over a type of
relations, which is not a proposition.
%
One possible solution is actually quite simple, and it just involves truncating
the $\Sigma$ type above to force it to be in $\mProp_1$.
\vspace{-0.2em}
\begin{align*}
  & \_\sim_{\setoidU}\_ : \setoidU \to \setoidU \to \mProp_1 \\
  & (A , a) \sim_{\setoidU} (B , b) =
   \ \parallel \Sigma\ (R : A \to B \to \mProp)\ (\inUU\ a\ b\ R) \parallel
\end{align*}

@@

In the definition above, $p$ has type $\parallel \Sigma\ (R: \El\,A \to \El\,B \to
\mProp)\ (...) \parallel$. If the type wasn't propositionally truncated, we could define
$p \vdash x \sim_{\El} y$ by extracting the relation out of the first component
of $p$, and apply it to $x, y$. That is, $p \vdash x \sim_{\El} y :\equiv
\pi_1\,p\,x\,y$. This would make the definition of $\_\sim_\setoidU\_$ and
$\_\vdash\_\sim_{\El}\_$ in line with how we defined $\setoidU$ and $\El$.

However, this doesn't work in our case, since the type of $p$ \emph{is}
propositionally truncated, hence it cannot be eliminated to construct a
proof-relevant object.
%
Fortunately, we can work around this limitation by defining $p \vdash x
\sim_{\El} y$ by induction on the codes $A\ B : \setoidU$, in a way that ends
up being logically equivalent to the proposition we would have obtained by
$\pi_1\ p\ x\ y$ if there were no truncation.
%
More precisely, we need to construct proofs that for any concrete $R$ and
$\mathsf{inR}$, the types $| (R, \mathsf{inR}) | \vdash x \sim_{\El} y$ and $R\ x\ y$
are logically equivalent. These in turn need to be defined mutually with
$\_\vdash\_\sim_{\El}\_$. We direct the interested reader to the Agda
formalization for the full details of these definitions, as they are quite
involved.

The full definition of the universe is concluded with the remaining definitions,
like $\reflu, \reflel$, etc., which can be adapted from their IR counterparts
more or less straightforwardly. The final result does not use
induction-recursion, but it's nevertheless an instance of infinitary
induction-induction. The ability to define arbitrary, infinitary
inductive-inductive types clashes, again, with our objective of keeping the
metatheory as small and simple as possible. The next step is therefore to reduce
this inductive-inductive universe to one that does not require (infinitary)
induction-induction.

@@

The reduction described in the previous section works generically for an
arbitrary finitary inductive-inductive type, thus giving a systematic way to
reduce finitary inductive-inductive definitions to inductive families. However,
it is not clear whether this method extends to \emph{infinitary}
induction-induction, of which the setoid universe defined in
Section~\ref{ii-universe} is an instance.
%
Of course, the absence of a general reduction method does not mean that we can't
reduce particular concrete instances of infinitary induction-induction, which is
exactly what we hope for our universe construction.

@@

Having $\mProp$ and a strong transport principle does help to some extent.
%
However, we would still need extensionality to derive the general eliminators
for $\inU$ and $\inUU$. In fact, as explained in the previous section,
to derive the general recursive-recursive eliminators we need to prove that the
corresponding graph relations are functional, which can't be done without funext.

>>>>

For convenience, we slightly deviate from $\mlttp$ both in the paper and in the
formalization, for instance by relying on pattern matching instead of
eliminators, and using primitive versions of $\mProp$-valued $\Pi$ and $\Sigma$
types instead of deriving them from truncation. We operate under the assumption
that everything can be equivalently carried out in $\mlttp$, although we have not
fully checked all the necessary details.

@@

We can show that the setoid model validates the usual basic type formers ($\Pi,
\Sigma$, etc.), function extensionality and a universe of strict
propositions with propositional extensionality \cite{setoid99}. Note that
we do not need identity types or inductive types (W-types) for this.

@@

Other design choices on the setoid universe are less essential, but still
require careful consideration. For instance, one question is whether the setoid
universe should support universe induction, thus exposing the inductive
structure of the codes. Such an elimination principle is known to be inconsistent
with univalence, although this is not an issue in our case; nevertheless it is
not immediately clear if the elimination principle can be justified by the
semantics, that is, if our encoding of the setoid universe in the model allows
to define such a universe eliminator.
%
The question arises because our final encoding of the setoid universe only
supports a weak form of elimination, for reasons that are explained in
Section~\ref{inductive-universe}. Although not currently needed, a stronger
eliminator might be necessary to justify universe induction. This problem should
not arise in the other encodings of the setoid universe (as given in
Section~\ref{ir-setoid-universe} and Section~\ref{ii-universe}).

@@

In the next section we transform our current inductive-recursive definition to
one that does not use induction-recursion.
%
The way this is done is inspired by the well-known trick to eliminate
induction-recursion described in Section~\ref{ir-universes}, but modified in a
novel way to account for the presence of $\mProp$-valued types. To our
knowledge, this is the first time this reduction method is applied to an
inductive-recursive type of this kind.

@@

Unfortunately this definition only manages to capture the idea, but does not
actually typecheck. In fact, $\_\sim_{\setoidU}\_$ should be a $\mProp_1$-valued
relation, so $A \sim_{\setoidU} B$ should be a proposition. However, the
$\Sigma$ type shown above clearly is not, since it quantifies over a type of
relations, which is not a proposition.
%
One possible solution is actually quite simple, and it just involves truncating
the $\Sigma$ type above to force it to be in $\mProp_1$.
\vspace{-0.2em}
\begin{align*}
  & \_\sim_{\setoidU}\_ : \setoidU \to \setoidU \to \mProp_1 \\
  & (A , a) \sim_{\setoidU} (B , b) =
   \ \parallel \Sigma\ (R : A \to B \to \mProp)\ (\inUU\ a\ b\ R) \parallel
\end{align*}

@@

In the definition above, $p$ has type $\parallel \Sigma\ (R: \El\,A \to \El\,B \to
\mProp)\ (...) \parallel$. If the type was not propositionally truncated, we could define
$p \vdash x \sim_{\El} y$ by extracting the relation out of the first component
of $p$, and apply it to $x, y$. That is, $p \vdash x \sim_{\El} y :\equiv
\pi_1\,p\,x\,y$. This would make the definition of $\_\sim_\setoidU\_$ and
$\_\vdash\_\sim_{\El}\_$ in line with how we defined $\setoidU$ and $\El$.

However, this does not work in our case, since the type of $p$ \emph{is}
propositionally truncated, hence it cannot be eliminated to construct a
proof-relevant object.
%
Fortunately, we can work around this limitation by defining $p \vdash x
\sim_{\El} y$ by induction on the codes $A\ B : \setoidU$, in a way that ends
up being logically equivalent to the proposition we would have obtained by
$\pi_1\ p\ x\ y$ if there were no truncation.
%
More precisely, we need to construct proofs that for any concrete $R$ and
$\mathsf{inR}$, the types $| (R, \mathsf{inR}) | \vdash x \sim_{\El} y$ and $R\ x\ y$
are logically equivalent. These in turn need to be defined mutually with
$\_\vdash\_\sim_{\El}\_$. We direct the interested reader to the Agda
formalization for the full details of these definitions, as they are quite
involved.

The full definition of the universe is concluded with the remaining definitions,
like $\reflu, \reflel$, etc., which can be adapted from their IR counterparts
more or less straightforwardly. The final result does not use
induction-recursion, but it is nevertheless an instance of infinitary
induction-induction. The ability to define arbitrary, infinitary
inductive-inductive types clashes, again, with our objective of keeping the
metatheory as small and simple as possible. The next step is therefore to reduce
this inductive-inductive universe to one that does not require (infinitary)
induction-induction.

@@

The reduction described in the previous section works generically for an
arbitrary finitary inductive-inductive type, thus giving a systematic way to
reduce finitary inductive-inductive definitions to inductive families. However,
it is not clear whether this method extends to \emph{infinitary}
induction-induction, of which the setoid universe defined in
Section~\ref{ii-universe} is an instance.
%
Of course, the absence of a general reduction method does not mean that we cannot
reduce particular concrete instances of infinitary induction-induction, which is
exactly what we hope for our universe construction.

@@

Having $\mProp$ and a strong transport principle does help to some extent.
%
However, we would still need extensionality to derive the general eliminators
for $\inU$ and $\inUU$. In fact, as explained in the previous section,
to derive the general recursive-recursive eliminators we need to prove that the
corresponding graph relations are functional, which cannot be done without funext.
